Rates of true polar wander in convecting planets

### 2015-03-18 tpw_rate.tex
Disorganized work

In order to estimate $\theta$ and $\lambda_i$ we must examine the linear momentume equation.

@@

As mantle convection and rotational dynamics of planetary bodies are usually considered separately, some extra care must be taken to establish self-consistent governing equations.  
Here we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.  The equations for mass, momentum, and energy then read

@@

Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure. 
Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires \_\_\_\_? nondimensional numbers to characterize it.

@@

Convenient choices for these numbers are shown in Table \ref{nondim} along with approximate Earthlike values.
In addition to the more usual Rayleigh and Prandtl numbers we include a Froude number, characterizing the ratio of centrifugal to gravitational forces, and $\Gamma$, a nondimensional characterization of density variations.
Nondimensionalizing with these parameters we obtain
\begin{equation}
\begin{aligned}
 - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\bf \Omega \times \Omega \times r} = 0
\end{aligned}
\end{equation}

@@

\begin{table}
\centering
\caption{Nondimensional numbers with approximate Earthlike values}
\label{nondim}
\begin{tabular}{@{}lcccc}
Symbol &  Number & Definition & Approximate value \\
\hline
Ra & Rayleigh &  $\rho_0 g_0 \alpha \Delta T R^3/\eta \kappa$ & $10^7$\\
Pr & Prandtl & $\eta/\rho_0 \kappa$ & $10^{23}$ \\
Fr & Froude & $\Omega_0^2 R/g_0$ & $10^{-3}$ \\
$\Gamma$ & Density deficit & $\alpha \Delta T$ & $10^{-2}$ \\
A & Aspect ratio & $R_i/R$ & $0.55$ \\
\end{tabular}
\end{table}

@@

\begin{equation}
-\int_V {\bf r} \times \nabla P^* + \int_V {\bf r} \times \nabla^2 {\bf v} - \mathrm{Ra} \int_V T {\bf r \times g} + \mathrm{Ra \; Fr} \int_V T {\bf r \times \Omega \times \Omega \times r} = 0
\end{equation}

@@

\begin{equation}
\begin{aligned}
\int_V T {\bf r \times \Omega \times \Omega \times r} = 0 \\
\end{aligned}
\end{equation}

\section{Rotational dynamics}
\label{sec:rotation}

\begin{equation}
{\bf \Omega \times C \cdot \Omega} = 0
\end{equation}

Following \citet{ricard1993polar}, the total moment of inertia of the mantle may be separated into three parts, a spherically symmetric contribution, a rotationally deformed contribution, and a contribution due to the internal mass anomalies:

\begin{equation}
C_{ij} = I_0 \delta_{ij} + J_{ij} + E_{ij}
\end{equation}

where ${\bf E} = \left[ \delta(t) + k^L(t) \right] * {\bf C}$ is the convective part convolved with the surface response ot internal density anomalies, and

We may write the nonlinear Liouville equations 
\begin{equation}
{\bf \Omega} \times {\bf \dot{\Omega} } = \frac{1}{\epsilon I_0 \tau} {\bf \Omega} \times \left( {\bf E \cdot \Omega} \right)
\label{eq:liouville}
\end{equation}

introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{\epsilon I_0 \tau} \left[ {\bf E \cdot \mitbf{\omega}} - \left( {\bf \mitbf{\omega} \cdot E \cdot \mitbf{\omega} } \right) \mitbf{\omega} \right]
\end{equation}

Note that the quantity in brackets is identical in form to the shear stress on a plane in classical elastostatics.
If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):

And here is the tedious algebra.  Let the prefactors $\frac{1}{\epsilon I_0 \tau}$ be denoted by $A$, and the pole of the coordinate system ($\theta = 0$) associated with the eigenvector for $\lambda_3$.  Therefore we find that $\mitbf{\omega} = \left[ \sin{\theta} \cos{\phi} \;\; \sin{\theta} \sin{\phi} \;\; \cos{\theta} \right]^T$.  Plugging this in, we find a system of three equations, which we want to solve for $\dot{\theta}$ and $\dot{\phi}$:

\begin{equation}
\begin{aligned}
\dot{\theta} &= - \frac{1}{2 \epsilon I_0 \tau} \sin{2 \theta} \left[ (\lambda_3-\lambda_1) \cos^2{\phi} + (\lambda_3-\lambda_2) \sin^2{\phi} \right] \\
\dot{\phi} &= \;\; \frac{1}{2 \epsilon I_0 \tau} \sin{2 \phi} \; (\lambda_2 - \lambda_1)
\end{aligned}
\label{milankovitch}
\end{equation}

Without loss of generality we may choose the coordinate system such that $\phi=0$:
\begin{equation}
\dot{\theta} = - \frac{1}{2 \epsilon \tau}\sin{2 \theta} \frac{(\lambda_3-\lambda_1)}{I_0}
\label{simple_milankovitch}
\end{equation}

These equations are a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}.  
Written this way, it is clear that the important quantities are $\theta$ and the differences between the eigenvalues of the convective moment $\bf E$.
It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
With this we may rewrite this in terms of the nondimensional numbers defined above, where 
$\epsilon \sim \mathrm{Fr}$ and $\tau \sim \eta / \rho g R = \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}$.

@@

The quantites $\Lambda_{ij}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
We consider relatively slowly rotating bodies here ($\mathrm{Fr} \ll 1$), so that the rotation does not affect the style of convection.  
In this regime, therefore, we neglect the dependence on $\mathrm{Fr}$.

@@

The nonhydrostatic moment of inertia is defined
\begin{equation}
J_{ij} = \int_{V_S} \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) 
\end{equation}

where $V_S$ is the reference spherical volume of the mantle.
If we substitute in Equation \ref{eos} we find

\begin{equation}
J_{ij} = I_0 \delta_{ij} + \int_{V_S} \rho_0 \alpha T \left( r_q r_q \delta_{ij} - r_i r_j \right) = I_0 \delta_{ij} + E_{ij} 
\end{equation}

@@

\begin{enumerate}
\item Figure out the best way of nondimensionalizing ($\alpha \Delta T$)
\item Connect nondimensionalization with Equation \ref{eq:liouville} via correspondence principle
\item Make notation everywhere consistent
\item More discussion of Davis+Kahan?
\item Make better figures
\item Integrate numerical results
\item Make the argument about degree-two scaling much sharper
\item Can I write equation \ref{eq:liouville} in terms of \ref{maccullagh}?
\item Discuss elastic lithosphere, Creveling
\item Discuss high-viscosity lower mantle
\item Earth-like parameters
\end{enumerate}

>>>>

In order to estimate $\theta$ and $\lambda_i$ we must examine the linear momentum equation.

@@

As mantle convection and rotational dynamics of planetary bodies are usually considered separately, some extra care must be taken to establish self-consistent governing equations.  
For simplicty we consider an isoviscous planet in a rotating reference frame with no internal heating in the incompressible Boussinesq approximation.  The equations for mass, momentum, and energy then read

@@

Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure.
Dimensional analysis of this system (cf. \citet{barenblatt1996scaling}) requires three nondimensional numbers to characterize it.
Convenient choices for these numbers are listed in Table \ref{nondim}, along with approximate Earth-like values for them.

In particular, the most important numbers are the Rayleigh number, characterizing the vigor of convection, and the ratio of centrifugal to gravitational forces.
This second nondimensional number does not have a uniformly agreed-upon name, but has been called a Froude number in analogy with other applications of inertial-to-gravitational effects \citep{mckenzie1968influence}, which we adopt here.
Since we have begun with equations that do not have inertia or compressibility, we have implicitly thrown out the dependence on the nondimensional numbers which characterize those effects (e.g., the Prandtl and dissipation numbers).
It would be straightforward to include them, but does not affect the overall treatment of the effects of rotation.

@@

Nondimensionalizing with these parameters we obtain
\begin{equation}
\begin{aligned}
 - \nabla P^* + \; \nabla^2{\bf v} - \mathrm{Ra} \; T \; {\bf g} + \mathrm{Ra \; Fr}\; T \;{\mitbf{\omega} \times \mitbf{\omega} \times r} = 0
\end{aligned}
\end{equation}

@@

\begin{table}
\centering
\caption{Nondimensional numbers with approximate Earthlike values}
\label{nondim}
\begin{tabular}{@{}lcccc}
Symbol &  Number & Definition & Approximate value \\
\hline
Ra & Rayleigh &  $\rho_0 g_0 \alpha \Delta T R^3/\eta \kappa$ & $10^7$\\
Fr & Froude & $\Omega_0^2 R/g_0$ & $10^{-3}$ \\
A & Aspect ratio & $R_i/R$ & $0.54$ \\
\end{tabular}
\end{table}

@@

\begin{equation}
-\int_V {\bf r} \times \nabla P^* + \int_V {\bf r} \times \nabla^2 {\bf v} - \mathrm{Ra} \int_V T {\bf r \times g} + \mathrm{Ra \; Fr} \int_V T {\bf r \times \mitbf{\omega} \times \mitbf{\omega} \times r} = 0
\end{equation}

@@

\begin{equation}
\begin{aligned}
\int_V T {\bf r \times \mitbf{\omega} \times \mitbf{\omega} \times r} = 0 \\
\end{aligned}
\end{equation}

It is useful to define a nondimensional eigenvalue difference (or ``eigengap'') $\Lambda_{ij} \equiv (\lambda_i - \lambda_j)/I_0$.  
This quantity represents the size of fluctuations in the convective moment of inertia compared to the reference value.
We also define a new nondimensional number (which will come in handy in intermediate steps) corresponding to the fractional density changes due to temperature variations: $\Gamma = \alpha \Delta T$.
With this we may rewrite Equation \ref{simple_milankovitch} in terms of the nondimensional numbers defined above.
The difference in the polar and equatorial moments of the hydrostatic planet $(C-A)$ is proportional to the rotational to gravitational forces, or Froude number \citep[e.g.][]{??}.
\begin{equation}
(C-A) \sim I_0 \mathrm{Fr}
\end{equation}

The time constant $T_1$ is related to the viscous tidal relaxation time of the planet. It is frequently represented as a weighted average of the different relaxation modes of the planet \citep[e.g.][]{ricard1993polar, greff2004upwelling}, but for our purposes it is enough to approximate it as a single mode:
\begin{equation}
T_1 \sim \eta / \rho g R = \frac{R^2}{\kappa} \Gamma/\mathrm{Ra}
\end{equation}

@@

The quantites $\Lambda_{ij}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
We consider relatively slowly rotating bodies here ($\mathrm{Fr} \ll 1$), and so make the simplifying assumption that the rotation does not significantly affect the style of convection.  
In this regime, therefore, we neglect the dependence on $\mathrm{Fr}$, and look for scalings of the form $\Lambda_{ij}(\mathrm{Ra})$ and $\theta(\mathrm{Ra})$.

@@

TODO: Needs work with notational consistency/connecting to previous scaling...

Fluctuations in the nonhydrostatic moment of inertia are due to temperature fluctuations in the interior and their spatial structure.

\begin{equation}
C_{ij} = \int_{V_S} \rho_0 \alpha T \left( r_q r_q \delta_{ij} - r_i r_j \right) 
\end{equation}
where $V_S$ is the reference spherical volume of the mantle.

These internal density loads may be convolved with surface responses to get the convective moment of inertia $\bf E$.

@@

\begin{enumerate}
\item Make notation everywhere consistent
\item More discussion of Davis+Kahan?
\item Make better figures
\item Integrate numerical results
\item Make the argument about degree-two scaling much sharper
\item Discuss elastic lithosphere, Creveling
\item Discuss high-viscosity lower mantle
\item Earth-like parameters
\end{enumerate}
